Modelling the Retinotectal projection - can competition provide a stopping mechanism?

### 2021-11-19 paper/paper.tex
Adds a figure showing expressions for the different manipulations
Ready to move on to the competition section in the 1st draft

Results are shown in Fig.\,\ref{f:ch}. Fig.\,\ref{f:ch}A indicates the colour
code used for retinal cell position; dorsal retinal cells are indicated by
green, nasal cells by red. Lightness gives position along the relevant axis so
that cells at the dorsal-nasal corner are yellow and those at the
ventral-transverse corner are black. Fig.\,\ref{f:ch}E shows the expected
layout of retinal ganglion cell axons on the tectum. The axons retain the
colour of their originating cell on the retina, as shown in
Fig.\,\ref{f:ch}A. Nasal cell axons should be arranged caudally; dorsal axons
laterally; axons should be evenly spaced and the pattern should span the
entire tectum.  Fig.\,\ref{f:ch}B--D show `fish net' plots of the axon
centroids at three timepoints. An axon's centroid is the average location of
its 4 branches. In these plots, lines are drawn between axons that are
expected to arrange themselves adjacently to one another by the end of the
simulation. The number of line crossings in this plot gives an indication of
the disorder in the arrangement. Fig.\,\ref{f:ch}G shows the positions of the
individual branches (i.e.~growth cones) at $t=300$ that contribute to the the
plot in Fig.\,\ref{f:ch}D. By this time, the topological order matches that
shown in Fig.\,\ref{f:ch}E. Opposing gradients have sorted the axons along the
two axes to form a grid of cells. However, the arrangement does not span the
full tectum. Fig.\,\ref{f:ch}F shows two pattern metrics; the sum of squared
distance errors of the axon centroids and the number of crossings in the fish
net plot of Fig.\,\ref{f:ch}D. The expected number of crossings is 0, as in
Fig.\,\ref{f:ch}E. Fig.\,\ref{f:ch}H shows selected axons (one from the centre
and one from each corner of the retina) and their position history.

@@

The result at $t=300$ is shown in Fig.\,\ref{f:chalt}A and B. The axon
centroids are arranged across the full tectal surface.  The arrangement is
slightly disordered around the middle of the tectum. This is caused by
numerical artefacts from the discretization of the ligand expression and the
fact that in the centre, the difference between the opposing receptor-ligand
signals is very small. Nevertheless, it is demonstrated that a modified form
for the tectal ligand expression (which was found easily by hand tuning)
allows a well ordered map to form for the wildtype system. Fig
.\,\ref{f:chalt}C shows the error metric, which is closer to 0 with the new
ligand expression.

@@

The `chemoaffinity only' model begins to encounter problems when we introduce
simulations of experimental manipulations. Fig\,\ref{f:chsurg} shows the
results of 6 simulated surgical manipulations. In row A, a patch of tectal
tissue has been cut out and rotated by a quarter turn. This is achieved in
simulation by rotating a square region of the tectal region

% Visualisation of ligand expression and gradients on the tectum for the
% graft-rotate 90 degrees simulated surgical manipulation
\begin{figure}
\includegraphics[width=\linewidth]{./images/Tissuevisb.png}
\caption{Visualisation of ligand expression and gradients on the tectum for the
graft-rotate 90 degrees simulated surgical manipulation}
\label{f:tvrot90}
\end{figure}

% The chemo manipulations figure is made up from 4 figures generated with
% graph_layout::g
%
% ./build/sim/agent/agent1 configs/simpler/m_eE_G.json  configs/simpler/e_tecrot90_fig4row1.json
% ./build/sim/agent/agent1 configs/simpler/m_eE_G.json  configs/simpler/e_tecrot180_fig4row2.json
% ./build/sim/agent/agent1 configs/simpler/m_eE_G.json  configs/simpler/e_tecswap_fig4row3.json
% ./build/sim/agent/agent1 configs/simpler/m_eE_G.json  configs/simpler/e_retablat_fig4row4.json
% ./build/sim/agent/agent1 configs/simpler/m_eE_G.json  configs/simpler/e_tecablat_fig4row5.json
% ./build/sim/agent/agent1 configs/simpler/m_eE_G.json  configs/simpler/e_mismatch_fig4row6.json
%
\begin{figure}
\includegraphics[width=0.8\linewidth]{./images/fig_chemo_manipulations.png}
\caption{Surgical manipulations in the chemoaffinity-only model. Row A: 90 degree
rotation. Row B: 180 degree rotation. Row C: Graft swap. Row D: mismatch ablation}
\label{f:chsurg}
\end{figure}

@@

To enable a simulation it is necessary to make assumptions about the nature of
the receptor-ligand signalled interactions (i.e. whether they are attractive
or repulsive) and about the pattern of ligand expression on the tectal
surface. We assume that i) All receptor-ligand signalled interactions are
repulsive, as for EphA-ephrinA
coupling \citep{drescher_vitro_1995,nakamoto_topographically_1996}; and ii)
all tectal ligand expressions vary exponentially along a given axis, following
Eq.\,\ref{e:tecligexp}. The results of running such a system are presented in
Fig.\,\ref{f:wt}.
%
Fig.\,\ref{f:wt}Ai shows a dual-colour map of retinal ganglion cell positions
on the square retina, with red encoding position on the $x$ axis (temporal to
nasal) and green encoding position on the $y$ axis (ventral to
dorsal). Fig.\,\ref{f:wt}Aii presents a fishnet plot of the expected location on
the tectal surface of each RGC at the end of the development process. Thus,
the `red' RGC from the ventro-nasal corner should map to the caudo-medial
corner of the tectum and the `green' RGC from the dorso-temporal corner should
find its way to the rostro-lateral corner on the tectum. The lines between the
dots of Fig.\,\ref{f:wt}Aii indicate the adjacency relationships between
cells. Thus, colour in each panel indicates a branch or axon's location of
origin on the retina.
%
Fig.\,\ref{f:wt}B shows the initial state of the system, with branches
stochastically arranged around the rostral edge of the tectum. Each dot
represents the centroid of $n=8$ branches per RGC axon. Figs.\,\ref{f:wt}C and \ref{f:wt}D
show the development of the system with time, with Fig.\,\ref{f:wt}D showing a
well ordered arrangement, with a majority of adjacency relationships matching
the experimental prediction (i.e. there are few crossed lines). The
intermediate timepoint at $t=40$ indicates that branches quickly cover the simulated
tectal surface, with final adjacency relationships developing over a longer
timescale. This dynamic behaviour is also evident in Fig.\,\ref{f:wt}E, which
shows \textbf{RMS error or some metric} of the axon centroids with respect to
the experimental prediction. Fig.\,\ref{f:wt}F shows the final
location of each branch in the simulation, which gives a more continuous view
of the order present in Fig.\,\ref{f:wt}D. Finally, Fig.\,\ref{f:wt}G shows the
branches and their centroid path history for five selected RGCs, with path
histories strongly resembling those presented by \citet{simpson_simple_2011}.

>>>>

Results are shown in Fig.\,\ref{f:ch}. Fig.\,\ref{f:ch}A indicates the colour
code used for retinal cell position; dorsal retinal cells are indicated by
green, nasal cells by red. Lightness gives position along the relevant axis so
that cells at the dorsal-nasal corner are yellow and those at the
ventral-transverse corner are black. Fig.\,\ref{f:ch}E shows the expected
layout of retinal ganglion cell axons on the tectum. The axons retain the
colour of their originating cell on the retina, as shown in
Fig.\,\ref{f:ch}A. Nasal cell axons should be arranged caudally; dorsal axons
laterally; axons should be evenly spaced and the pattern should span the
entire tectum.  Fig.\,\ref{f:ch}B--D show `fish net' plots of the axon
centroids at three timepoints. An axon's centroid is the average location of
its 4 branches. In these plots, lines are drawn between axons that are
expected to arrange themselves adjacently to one another by the end of the
simulation. The number of line crossings in this plot gives an indication of
the disorder in the arrangement. Fig.\,\ref{f:ch}G shows the positions of the
individual branches (i.e.~growth cones) at $t=300$ that contribute to the the
plot in Fig.\,\ref{f:ch}D. By this time, the topological order matches that
shown in Fig.\,\ref{f:ch}E. Opposing gradients have sorted the axons along the
two axes to form a grid of cells. However, the arrangement does not span the
full tectum. Fig.\,\ref{f:ch}F shows two pattern metrics. The first metric,
$\epsilon$, is the sum of squared distance errors of the axon centroids, with
$\mathbf{x}_{k}$, the location of the centroid of the branches of axon $k$ and
$\mathbf{x}'_{k}$ the target tectal location for axon $k$:
%
\begin{equation}\label{e:eps}
\epsilon = \sum_k |\mathbf{x}_{k} - \mathbf{x}'_{k}|^2.
\end{equation}
%
$\epsilon$ tends to 0 as the pattern approaches the target layout. However,
two patterns may have the same non-zero value for $\epsilon$, despite one
being qualitatively more disordered than the other. For this reason, we also
use the number of crossings in the fish net plot of Fig.\,\ref{f:ch}D as a
second metric, called $\eta$. The expected number of crossings in the wildtype
pattern is 0, as in Fig.\,\ref{f:ch}E, although note that in some patterns,
the expected number of crossings is non-zero. Fig.\,\ref{f:ch}H shows selected
axons (one from the centre and one from each corner of the retina) and their
position history.

@@

The result at $t=300$ is shown in Figs.\,\ref{f:chalt}A and \ref{f:chalt}B. The
axon centroids are arranged across the full tectal surface.  The arrangement
is slightly disordered around the middle of the tectum. This is caused by
numerical artefacts from the discretization of the ligand expression and the
fact that in the centre, the difference between the opposing receptor-ligand
signals is very small. Nevertheless, the result demonstrates that a modified
form for the tectal ligand expression (which was found easily by hand tuning)
allows a well ordered map to form for the wildtype system. Fig
.\,\ref{f:chalt}C shows the error metric (\textbf{FIXIMAGE}), $\epsilon$,
which is much closer to 0 with the new ligand expression.

@@

This `chemoaffinity only' model begins to encounter problems when we introduce
simulations of experimental surgical manipulations. Simulated surgery is
performed by manipulating the maps shown in
Fig.\,\ref{f:ex}. Fig.\,\ref{f:trot90} shows the expression maps and the
resulting gradients that result from the surgical manipulation in which a
graft of the tectum is rotated by a quarter turn and replaced. We rotate an
8x8 square of the tectum and then recompute the gradients. As the ligand
expression is discretize rather than continuous, we compute the gradient
numerically, with the gradient in direction $x$ computed as the difference
between the expression of the two neighbouring elements divided by twice the
element distance. As there are four ligands expressed on the tectum, there are
four pairs of gradient maps, each of which is shown in Fig.\,\ref{f:trot90}.

% Visualisation of ligand expression and gradients on the tectum for the
% graft-rotate 90 degrees simulated surgical manipulation
% ./build/sim/agent/tissuevis configs/simpler/m_eE_G.json configs/simpler/e_tecrot90_fig4row1.json
%
\begin{figure}
\includegraphics[width=\linewidth]{./images/Tissuevisb.png}
\caption{Visualisation of ligand expression and gradients on the tectum for the
graft-rotate 90 degrees simulated surgical manipulation}
\label{f:trot90}
\end{figure}

Fig.\,\ref{f:tmanip} shows the tectal ligand expression for ligands $L_1$ and
$L_2$ for the five other surgical manipulations applied in this work. These
are a 180\degree~rotation (Fig\,\ref{f:tmanip}A), a graft-swap manipulation
(Fig\,\ref{f:tmanip}B), ablation of half the retina (Fig\,\ref{f:tmanip}C),
and ablation of half the tectum (Fig\,\ref{f:tmanip}D). A mismatch
manipulation was also performed in which which half of the retina is removed
along with the half of the tectum to which the remaining retina would project.

\begin{figure}
\includegraphics[width=\linewidth]{./images/expressions_manipulations.png}
\caption{Visualisation of ligand expression and gradients on the tectum for the
other simulated surgical manipulations}
\label{f:tmanip}
\end{figure}

Fig\,\ref{f:chsurg} shows the results of 6 simulated surgical
manipulations. In row A, a patch of tectal tissue has been cut out and rotated
by a quarter turn, as in Fig.\,\ref{f:trot90}. From left to right, graphs show
i) the expected final layout of the tectum, according to the results of
experimental work, with the colour code giving the origin location of the
axon's soma on the retina, ii) The locations of individual axon branches at
simulation end, iii) the centroids of each axons' branches and iv) a rather
useless selected axons box \textbf{FIXME}. Simulation time was 1000 time steps. The final
locations of branches and axon centroids indicates that a qualitative rotation
of the graft patch has occurred, with red branches inside the patch being
located at the lateral-caudal corner rather than the medial-caudal corner. A
portion of the graft region (the medial side, after rotation) is free of
branches, indicating that the pattern is partial. A similar result is seen for
the 180\degree~example in row B, in whcih the caudal half of the rotated patch
is free of branches, but the rostral half contains branches with the correctly
rotated orientations.

Fig.\,\ref{f:chsurg}C shows the result of the patch swap manipulation, in
which 12x4 element grafts, one from the caudal side of the tectum and one from
the rostral side are swapped. This simulation is unsuccessful in patterning
the caudal half of the tectum. The large gradients, both positive and
negative, caused by the discontinuities along the long edges of the grafts
attracts almost all the branches (whose initial position is the rostral edge)
and very few pass through to the caudal side.

The retinal ablation shown in Fig.\,\ref{f:chsurg}D is also unsuccessful, with
the axons from the surviving nasal retina (the transverse half was removed)
failing to arrange themselves across the entire tectum, due to the pure
`position specification' which the dual gradient model achieves. Similarly,
the tectal ablation simulation fails to reproduce experimentally observed
results in which the retinal map `squashes up' within the remaining tectal
region. Instead, axons from the transverse retina arrange with normal spacing
and axons from the nasal retina are pushd to the caudal-most side of the
remaining tectum (Fig.\,\ref{f:chsurg}E). A similar result occurs in the
mismatch experiment (Fig.\,\ref{f:chsurg}F), but in this case there are no
axons from teh transverse retina, which was been ablated and only the axons
from the nasal retina are present, arranged in a line.

% The chemo manipulations figure is made up from 4 figures generated with
% graph_layout::g
%
% ./build/sim/agent/agent1 configs/simpler/m_eE_G.json  configs/simpler/e_tecrot90_fig4row1.json
% ./build/sim/agent/agent1 configs/simpler/m_eE_G.json  configs/simpler/e_tecrot180_fig4row2.json
% ./build/sim/agent/agent1 configs/simpler/m_eE_G.json  configs/simpler/e_tecswap_fig4row3.json
% ./build/sim/agent/agent1 configs/simpler/m_eE_G.json  configs/simpler/e_retablat_fig4row4.json
% ./build/sim/agent/agent1 configs/simpler/m_eE_G.json  configs/simpler/e_tecablat_fig4row5.json
% ./build/sim/agent/agent1 configs/simpler/m_eE_G.json  configs/simpler/e_mismatch_fig4row6.json
%
\begin{figure}
\includegraphics[width=0.95\linewidth]{./images/fig_chemo_manipulations.png}
\caption{Surgical manipulations in the chemoaffinity-only model. \textbf{A} 90 degree
rotation. textbf{B} 180 degree rotation. \textbf{C} Graft swap. \textbf{D}
Retinal ablation \textbf{E} Tectal ablation \textbf{F} mismatch ablation}
\label{f:chsurg}
\end{figure}

Together, these results show that, while a pure gradient-following
implementation of Sperry's chemoaffinity theory is able to reproduce the
wildtype tectal arrangement of retinal ganglion cell axons, it cannot account
for the results of surgical manipulation experiments.

\subsection*{Competition}

\section{ZEROTH DRAFT/NOTES}

@@

%
Fig.\,\ref{f:wt}Ai shows a dual-colour map of retinal ganglion cell positions
on the square retina, with red encoding position on the $x$ axis (temporal to
nasal) and green encoding position on the $y$ axis (ventral to
dorsal). Fig.\,\ref{f:wt}Aii presents a fishnet plot of the expected location on
the tectal surface of each RGC at the end of the development process. Thus,
the `red' RGC from the ventro-nasal corner should map to the caudo-medial
corner of the tectum and the `green' RGC from the dorso-temporal corner should
find its way to the rostro-lateral corner on the tectum. The lines between the
dots of Fig.\,\ref{f:wt}Aii indicate the adjacency relationships between
cells. Thus, colour in each panel indicates a branch or axon's location of
origin on the retina.
%
Fig.\,\ref{f:wt}B shows the initial state of the system, with branches
stochastically arranged around the rostral edge of the tectum. Each dot
represents the centroid of $n=8$ branches per RGC axon. Figs.\,\ref{f:wt}C and \ref{f:wt}D
show the development of the system with time, with Fig.\,\ref{f:wt}D showing a
well ordered arrangement, with a majority of adjacency relationships matching
the experimental prediction (i.e. there are few crossed lines). The
intermediate timepoint at $t=40$ indicates that branches quickly cover the simulated
tectal surface, with final adjacency relationships developing over a longer
timescale. This dynamic behaviour is also evident in Fig.\,\ref{f:wt}E, which
shows \textbf{RMS error or some metric} of the axon centroids with respect to
the experimental prediction. Fig.\,\ref{f:wt}F shows the final
location of each branch in the simulation, which gives a more continuous view
of the order present in Fig.\,\ref{f:wt}D. Finally, Fig.\,\ref{f:wt}G shows the
branches and their centroid path history for five selected RGCs, with path
histories strongly resembling those presented by \citet{simpson_simple_2011}.
